---
title: "Utilizing NLCD Data - Revision"
author: "David Barbour"
date: "2025-07-05"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C:/Users/Dacat/OneDrive/NLCD data tutorial/Submit materials/Revision") #Set working directory with geocoded ames file here
```

Revision Changes 7/11/2025:
Created scalable boundary selection for 1000M buffer coverage to avoid edge cases and increase accuracy. This removes the need to download the entire eastern side of the county for larger buffer sizes. Clearing objects and reopening R may be required to change scaled boundary selection size. Additional checks for projection type have also been included to better explain the CRS matching. Limited repetitive transformations of projection type to reduce potential accuracy loss.






Please note:
rJava is required for OpenStreetMap satellite images, rJava also requires Java to be installed on your PC. This file was created with Version 8 Update 431 (Build 1.8.0_431-b10) installed on the host PC.

Skipping the OpenStreetMap portion will allow you avoid these packages and skip this step. While visualizing the area may be useful in analysis by giving an expectation of what the data should match to, it is not needed to extract the data for testing itself.

ggplot2 (for visualization) and dplyr (for data manipulation) are included as part of the tidyverse package.




SECTION 1. This section uses 100m Buffers and NLCD data from 2011 to explain the process step by step. Section 2 will break our data down by availability of data year to year for higher confidence in the accuracy of the data.





Load Required Libraries

This chunk installs and loads the necessary R libraries for working with National Land Cover Database (NLCD) data and performing spatial data analysis.
```{r}
#Uncomment the following if packages haven't been installed before
#install.packages(c("sf", "raster", "tigris", "FedData", "tidyverse", "exactextractr", "landscapemetrics"))
#Optional section
#install.packages(c("OpenStreetMap", "rJava"))


# Load required libraries:
library(sf)             # Simple Features (sf) for vector data processing
library(raster)         # For raster data manipulation
library(tigris)         # For downloading US Census geospatial data
library(FedData)        # Federal datasets, including NLCD
library(tidyverse)      # Collection of packages (ggplot2, dplyr, etc.)
library(exactextractr)  # Efficient raster extraction using polygons
# library(rJava)          # Java integration for OpenStreetMap usage - OPtional
# library(OpenStreetMap)  # Visualization and satellite imagery - Optional
```



Define Study Area

The first portion of the analysis collects needed data. It uses the tigris package and its places function tigris::places() to fetch boundaries for Iowa, before filtering down to Ames itself. Should the user wish to alter an alternative dataset, merely replace the location selection.


Something to note with the Tigris package is that it will default to pulling boundaries from 2022. Possible mismatch in boundaries could exist depending on time period and location due to shifting boundaries (Expansion, Annexation, Separation). Explicitly calling a year (as done above) may prove useful but be sure to cross reference available years with the census Bureau's TIGER/Line Shapefiles.

A constant throughout this process is to make sure that your data match CRS and Datum. Your EPSG code is essentially a projected tile on the earth while Longitude Latitude or Easting Northing combinations are points in those tiles. NAD83 / Conus Albers (EPSG:5070) is what we will use here, due to that being the native form of the NLCD files. If you're not sure why a mismatch may be an issue think of a Mercator projection and a 3D Spherical Map. You can't simply overlay them and expect them to match, they must be fitted one way or the other.







Tigris has multiple levels of data you can use. Pick a scale fitting for your area of interest (City, County, State).

In this section we pull the boundary files for Iowa state, then filter for the city of Ames. Then we expand our coverage with a bounding box, pull the NLCD data, set the boundary polygon to the proper CRS format (ESPG 5070) for overlay, and raster it. When in doubt always confirm your CRS when matching two files 70% of the errors we've encountered are due to mismatching CRS or accidentally clearing the CRS when forming a data frame.
```{r}
# Get place boundaries for Iowa. places(): Fetches the boundaries for all places (cities, towns, etc.) in Iowa.
cities <- places(state = "IA")

# Filter for a specific city (e.g., "Ames")
specific_city <- cities %>%
  filter(NAME == "Ames")
```

The default CRS for Tigris is ESPG 4269 we can confirm our projection, or if another source is used for the shapefile,  with st_crs ensure compatibility with the process. This will need to be reprojected into EPSG 5070 later (see code line 99) to avoid misalignment. It will not fundamentally change the underlying pairings between houses and the NLCD but will affect our boundary overlays, as long as the homes fall within the downloaded raster (which on this scale is a non issue, but worth mentioning for reapplication purposes). Note: this Rmd was created without files using alternative projections and so difficulties adapting another projection type may occur.
```{r}
st_crs(specific_city)
```




```{r}
# Transform to NAD83 / Conus Albers (EPSG:5070)
specific_city <- st_transform(specific_city, crs = 5070)

# City boundary plot
specific_city %>%
  sf::st_geometry() %>%
  plot(border = "darkblue", lwd = 2)

```



We'll also need extra space to avoid spillovers on our buffers later in the process. Here we expand the shapefile for the download area
```{r}
#Expand the bounding box of the specific_city polygon
expanded_bbox <- st_bbox(specific_city)
xpad <- diff(expanded_bbox[c("xmin", "xmax")]) * .2 #20% larger
ypad <- diff(expanded_bbox[c("ymin", "ymax")]) * .2

expanded_bbox[c("xmin", "xmax")] <- expanded_bbox[c("xmin", "xmax")] + c(-xpad, xpad)
expanded_bbox[c("ymin", "ymax")] <- expanded_bbox[c("ymin", "ymax")] + c(-ypad, ypad)

#Convert to sf object (this will be your NLCD download area)
specific_city_scaled <- st_as_sfc(expanded_bbox) %>%
  st_sf(crs = st_crs(specific_city))  # In EPSG:5070
```


```{r}
# Get NLCD data 
nlcd_data <- 
  FedData::get_nlcd(
    template = specific_city_scaled,
    year = 2011,
    label = "Ames"
  )
```

The get_nlcd command from the FedData package allows us to download NLCD data for our region of interest. Thus, FedData::get_nlcd: downloads NLCD (National Land Cover Database) raster data for 2011, clipped to Ames.

 Here we see the starting projection difference
```{r}
st_crs(nlcd_data)
```

By Rastering you're essentially changing lines and vectors into pixels or "cells" with data. Like a .SVG to .PNG or .JPEG
```{r, fig.width=12, fig.height=6}
# Save plot as a PNG
png("Figure_1.png", width = 12, height = 6, units = "in", res = 300)

# Plot NLCD raster and transformed Ames polygon
raster::plot(nlcd_data)

# Add city boundary to the plot
specific_city %>%
  sf::st_geometry() %>%
  plot(border = "darkblue", lwd = 2, add = TRUE)

# Close the PNG device
dev.off()
```
Optional step for extensions:

There are other data options with the NLCD that we won’t use here such as elevation, temperature, canopy, etc. For example below we fetch and plot the canopy data for Ames (effectively, the degree of leaf coverage) using the commands we have already developed before

```{r}
# Fetch NLCD canopy data for Ames
NLCD_canopy <- 
  FedData::get_nlcd(
    template = specific_city,
    year = 2011,
    label = "Ames Canopy",     # Descriptive label for the dataset
    dataset = "canopy"         # Dataset type (canopy/leaf coverage)
  )

# Plot the SpatRaster (NLCD canopy) directly
raster::plot(NLCD_canopy)

```

Now we bring in the Ames data set that includes longitude latitude data for houses in Ames. Our code first reads a CSV file containing geocoded data of houses and displays the first few observations for each variable.
```{r}
# Load the Ames dataset, which includes longitude and latitude data for houses
ames <- read.table("geocoded_ames_plus_schooldist.csv", 
                   header = TRUE,       # First row contains column names
                   sep = ",",           # Comma-separated values
                   stringsAsFactors = FALSE)  # Prevents automatic conversion of strings to factors
head(ames)  # Display the first few rows of the dataset
```

This here is a basic plot of the home in longitude and latitude.

Now we bring in the Ames data set that includes longitude latitude data for houses in Ames. We then convert the data into an sf object using the st_as_sf command, utilizing the information of the latitude and longitude of each house and plot the houses as a point layer. The gap in the middle of the map is the Iowa State University campus.

```{r}
# Convert the Ames dataset into a simple features (sf) object using longitude and latitude
ames_sf <- st_as_sf(ames, coords = c('Longitude', 'Latitude'))

# Create a basic map of the houses using ggplot2
ggplot(ames_sf) + 
  geom_sf() 

```


We can also create a nicer, more informative visualization of the house point plot. Below we show the point map again with the neighborhoods colored.

Our code below reduces neighborhood names to 10-character abbreviations for better clarity in plots. It then uses ggplot2 to visualize neighborhoods with distinct colors.
```{r}
#Reducing the size of the Neighborhood variable to 10 characters. This makes the legend more manageable in size.
ames_sf <- ames_sf %>% 
  mutate(Neighborhood_Abbr = substring(Neighborhood, 0, 10))

# Plot the map with neighborhoods distinguished by color
ggplot(ames_sf) + 
  geom_sf(aes(color = Neighborhood_Abbr)) +  # Color points by neighborhood
  labs(color = "Neighborhoods") +           # Label for the legend
  scale_fill_continuous(guide = guide_legend()) +  # Add a continuous legend
  theme(legend.position = "right",          # Position the legend to the right
        legend.direction = "vertical")      # Arrange legend items vertically

```


Satellite Visualization with OpenStreetMap

OpenStreetMap provides a satellite view of Ames, Iowa, allowing us to visually assess the spatial data in context. While this step isn’t necessary for our final analysis, it serves as a valuable tool to correlate the NLCD data points with true-color imagery.

Note: The rJava package and Java itself (installed on your system) are prerequisites for this step. Ensure compatibility between your Java installation and R.

Our code defines Ames's bounding box and fetches a basemap from OpenStreetMap. It reprojects the map to WGS84 CRS for compatibility. Then it overlays Ames's neighborhood boundaries on the map. This step helps you verify that your geographic data aligns accurately with the real-world context of Ames.

Openstreetmap is can be fairly unstable day to day and so we leave this commented out unless desired
```{r}
# Define the spatial extent for OpenStreetMap
#lat1 <- min(ames$Latitude)  # Minimum latitude
#lat2 <- max(ames$Latitude)  # Maximum latitude
#lon1 <- min(ames$Longitude) # Minimum longitude
#lon2 <- max(ames$Longitude) # Maximum longitude

#ames_map <- openmap(
#  c(lat2, lon1),  # Top-left corner (lat, lon)
#  c(lat1, lon2),  # Bottom-right corner (lat, lon)
#  zoom = NULL,    # Automatically determine zoom level
#  type = "bing",  # Use Bing Maps as the base layer
#  mergeTiles = TRUE  # Combine tiles into one map
#)

```

This still needs to be turned into a projection we can use.
```{r}
# reproject
#ames_map <- openproj(ames_map)

# Convert map into a ggplot-compatible object and add house points
#ames_map_plt <- OpenStreetMap::autoplot.OpenStreetMap(ames_map) + 
#  geom_sf(data = ames_sf,  # Add house locations from `ames_sf`
#          aes(x = ames$Longitude, y = ames$Latitude),
#          color = "red", size = 1) +  # Points are red for visibility
#  xlab("Longitude (°E)") + 
#  ylab("Latitude (°S)")

#ames_map_plt

```

This step helps you verify that your geographic data aligns accurately with the real-world context of Ames.

Here we take the Ames NLCD data and set the CRS on the ames_sf object. It's good to always double check the CRS or reset it to the correct value. During the creation of this paper it was found that making unrelated changes to the sf object can sometimes clear it. (Removing the code below will create an error for subsequent code, due to the need for a starting CRS before matching). 

It is important that we set the correct starting ESPG (in our case 4326) so that the transformation to match the nlcd data is not incorrectly Proccessed
```{r}
# Assigns CRS to match geocoded dataset baseline (EPSG:4326)
st_crs(ames_sf) <- 4326 
```



Let's attempt a combination of NLCD and the houses.
```{r, fig.width=14, fig.height=8}

# Transform neighborhoods to match the NLCD CRS if not already done
ames_sf_transformed <- st_transform(ames_sf, st_crs(nlcd_data)) 


# Plot NLCD raster
plot(nlcd_data,
     legend = TRUE,
     axes = TRUE)                    # Show Easting/Northing axes

# Add house points (small, styled dots)
plot(st_geometry(ames_sf_transformed),
     add = TRUE,
     pch = 21,                        # Circle points
     bg = "cyan",                    # Fill
     cex = 1)                       # Size

```
This is a bit far away so for future plots well use a bounding box from the original specific_city shape to zoom in
```{r, fig.width=14, fig.height=8}
# Get the bounding box of the original area
bbox <- st_bbox(specific_city)



png("Figure_2.png", width = 12, height = 6, units = "in", res = 300)

# Plot the NLCD raster but zoomed to original specific_city extent
plot(nlcd_data,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Re add dots
plot(st_geometry(ames_sf_transformed),
     add = TRUE,
     pch = 21,                       
     bg = "cyan",                    
     cex = 1)

dev.off()
```





*********************************************************** Note:
Prior version used ggplot for nicer visualizations, however due to issues with the geom_sf command reprojecting into ESPG:4326 (without affecting the underlying data projection). It will no longer be used for graphs where data and visual matching is important. This decision has been made to enhance visual fidelity of our spatial operations



***If the user would like to continue with ggplot we retained a baseline to get started. From converting the coordinate system to creating a dataframe for plotting in ggplot
Explicit extraction of the variables to enable our x y axis labeling in ggplot2
```{r}
# Extract Easting and Northing from geometry
#ames_coords <- ames_sf_transformed %>%
#  mutate(
#    Easting = st_coordinates(.)[, 1],
#    Northing = st_coordinates(.)[, 2]
#  )
```

*** plot with the extracted Easting and Northing values. You'll notice a few minor point shifts with geom_point compared to geom_sf, along with the lack of longitude latitude coordinates. Geom_sf would be required to represent the buffer shapes. Again this is what the underlying ESPG 5070 data values were all along it's merely ggplot2's reprojection that was out of place. It would not have impacted the data extraction but accurate visualization is important in it's own right, particularly when buffer overflow is possible.
```{r}
#ggplot(ames_coords, aes(x = Easting, y = Northing, color = Neighborhood_Abbr)) +
#  geom_point(size = 1.5) +
#  scale_color_viridis_d(name = "Neighborhood") +
#  labs(x = "Easting (m)", y = "Northing (m)") +
#  theme_minimal()
```

*** GGplot requires a data frame to work with, so the raster file must be interpreted into one.
```{r}
#Convert raster to data frame for ggplot2
#nlcd_df <- as.data.frame(nlcd_data, xy = TRUE) # The 'xy = TRUE' parameter ensures that the x (Easting) and y (Northing) coordinates of each raster cell are included as columns.

#Rename columns for clarity. The default column names here are x, y, and Class
#colnames(nlcd_df) <- c("Easting", "Northing", "Land_Cover")

#declare proper projection
#nlcd_df <- st_as_sf(nlcd_df, coords = c("Easting", "Northing"), crs = 5070)

```



```{r, fig.width=12, fig.height=8}
# Use fig.width and fig.height to adjust the figure dimensions in the R Markdown chunk to avoid crushing the data into too small of a space, and for better visualizing

# First plot: Ames neighborhoods overlaid on NLCD land cover
#ggplot() +
  # Add the NLCD raster layer with land cover categories
#  geom_raster(data = nlcd_df, aes(x = Easting, y = Northing, fill = Land_Cover)) +
#  scale_fill_viridis_d(name = "Land Cover") + # Use a discrete viridis color scale for land cover categories
  
#geom_point(data = ames_coords, aes(x = Easting, y = Northing, color = Neighborhood_Abbr)) +
#  scale_color_viridis_d(name = "Neighborhood") +
#  labs(x = "Easting (m)", y = "Northing (m)") +
#  theme_minimal()

# Second plot: Ames neighborhoods with a visible color overlay for better visibility, all other options identical, notice the home on the far right which blended in with the prior colors
#ggplot() +
#  geom_raster(data = nlcd_df, aes(x = Easting, y = Northing, fill = Land_Cover)) +
#  scale_fill_viridis_d(name = "Land Cover") + 
  
#geom_point(data = ames_coords, aes(x = Easting, y = Northing), color = "lightcoral") +
#  scale_color_viridis_d(name = "Neighborhood") +
#  labs(x = "Easting (m)", y = "Northing (m)") +
#  theme_minimal()
```



We conduct buffer analysis on the locations of each house in our dataset. We start by creating 100-meter radius buffer zones around houses using the st buffer function (this radius can of course be adjusted to specific contexts and any need for sensitivity analysis of results). We extract NLCD land cover data within each buffer using exactextractr. We then visualize buffers overlaid on NLCD data

```{r, fig.width=14, fig.height=8}

# This creates a buffer zone of 100 meters around each house feature in the spatial dataset, you may choose to use a greater or lesser buffer in your particular situation.
houses_buffer <- st_buffer(ames_sf_transformed, dist = 100)

# Extract land cover data values within the new buffer
# exact_extract retrieves raster values that intersect with each buffer zone
land_cover_raw <- exact_extract(nlcd_data, houses_buffer)


png("Figure_3.png", width = 12, height = 6, units = "in", res = 300)
# Plot the NLCD raster
plot(nlcd_data,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Add semi-transparent red buffers on top
plot(st_geometry(houses_buffer),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)

dev.off()
```

Transform the output into readable data:

The next step requires processing the extracted NLCD data for the buffer zones. We convert the information we have extracted into land cover proportions for each buffer. We then ``map'' NLCD numeric codes to descriptive land cover types. Finally, we filter and summarize the land cover data.
```{r}

# Convert the list of extracted values to a data frame
land_cover_df <- land_cover_raw %>%
  # Apply a function to each element in the list, converting it into a data frame with frequency counts
  lapply(function(x) as.data.frame(table(x$value))) %>%
  # Combine all individual data frames into one, adding an ID column for each buffer
  bind_rows(.id = "Buffer_ID") %>%
  # Group by Buffer_ID for calculations within each buffer
  group_by(Buffer_ID) %>%
  # Calculate the proportion of each land cover type within each buffer
  mutate(Proportion = Freq / sum(Freq))

# Create a mapping from NLCD class codes to descriptive land cover types
nlcd_mapping <- data.frame(
    Column = c("11", "12", "21", "22", "23", "24", 
               "31", "41", "42", "43", "51", 
               "52", "71", "72", "73", "74", 
               "81", "82", "90", "95"),  # NLCD land cover codes
    Land_Cover_Type = c("Open Water", "Perennial Ice/Snow",
                        "Developed, Open Space", "Developed, Low Intensity", 
                        "Developed, Medium Intensity", "Developed, High Intensity", 
                        "Barren Land", "Deciduous Forest", "Evergreen Forest", 
                        "Mixed Forest", "Dwarf Scrub", "Shrub/Scrub", 
                        "Grassland/Herbaceous", "Sedge/Herbaceous", 
                        "Lichens", "Moss", "Pasture/Hay", 
                        "Cultivated Crops", "Woody Wetlands", 
                        "Emergent Herbaceous Wetlands")  # Descriptive names based on the land cover database
)

# Create a lookup map for translating NLCD codes to land cover types
rename_map <- setNames(nlcd_mapping$Land_Cover_Type, nlcd_mapping$Column)

# Add a new column to the data frame for land cover types
land_cover_df <- land_cover_df %>%
  mutate(
    Land_Cover_Type = rename_map[as.character(Var1)]  # Map codes to land cover names
  )

# Check for any unmapped NLCD codes
unmapped_values <- land_cover_df %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values) > 0) {
  message("Unmapped land cover types detected: ", 
          paste(unique(unmapped_values$Var1), collapse = ", "))
}

# Optionally, filter out rows with NA land cover types to clean the data
land_cover_df_100m_example <- land_cover_df %>% filter(!is.na(Land_Cover_Type))

# Summarize the cleaned and mapped data
summary(land_cover_df_100m_example)

```

Next, we join the processed land cover data back to the house dataset (ames_sf_transformed). We then visualize the results, e.g., showing proportions of land cover types for a specific house buffer.
```{r}
# Pivot land_cover_df_100m_example to wide format and convert Buffer_ID to integer
land_cover_wide <- land_cover_df_100m_example %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%  # Select relevant columns for pivoting
  pivot_wider(
    names_from = Land_Cover_Type,  # Column names will be land cover types
    values_from = Proportion,  # Cell values will be the proportion of each land cover type
    values_fill = 0  # Fill missing land cover types with 0
  ) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))  # Convert Buffer_ID to integer for joining compatibility

# Check the updated wide-format dataframe to confirm the structure
head(land_cover_wide)

```

```{r}
# Join the wide land cover data to ames_sf_transformed using Buffer_ID
ames_sf_100m_example <- ames_sf_transformed %>%
  left_join(land_cover_wide, by = c("X" = "Buffer_ID"))  # Merge land cover data into spatial dataframe
ames_sf_tib <- as_tibble(ames_sf_100m_example)  # Convert the spatial dataframe into a tibble for easier inspection, this is purely for readability and not necessary

# View the updated spatial dataframe with land cover proportions added
head(ames_sf_tib)

```

How do we know it worked? Let's analyze house X 1 from the original data and buffer ID 1 from the added data.

This section verifies the accuracy of the land cover data integration by examining the proportions for a specific house (X == 1) and its corresponding buffer (Buffer_ID == 1).
```{r}
# Filter for the first house (X == 1)
house_1_stats <- land_cover_wide %>%
  filter(Buffer_ID == 1) %>%  # Filter to the first buffer (ID 1)
  pivot_longer(
    cols = -Buffer_ID,  # Select all columns except Buffer_ID for reshaping
    names_to = "Land_Cover_Type",  # Create a column for land cover types
    values_to = "Proportion"  # Create a column for proportions
  )

# View the stats
house_1_stats

# Bar plot of proportions for house 1
ggplot(house_1_stats, aes(x = Land_Cover_Type, y = Proportion, fill = Land_Cover_Type)) +
  geom_bar(stat = "identity") +  # Create a bar chart where bar heights represent proportions
  scale_fill_viridis_d(name = "Land Cover Type", guide = "none") +  # Discrete color scale
  labs(
    x = "Land Cover Type",  # X-axis label
    y = "Proportion"  # Y-axis label
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

# Save the plot
ggsave("Figure_4.png", width = 10, height = 6, dpi = 300)
```

So we should expect to see these proportions around house #1 with zero land cover that matches Barren land or Mixed Forrest as an example.
```{r, fig.width=14, fig.height=8}
# Filter for the buffer and overlay it on the map
house_1_buffer <- houses_buffer %>% filter(X == 1)  # Select buffer for house 1 using X == 1

# Plot the buffer and NLCD data
plot(nlcd_data,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

plot(st_geometry(house_1_buffer),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)
```

Well that's kind of hard to see so let's zoom in more
```{r}
# Get the bounding box of the buffer for house 1
house_1_bbox <- st_bbox(house_1_buffer)

# Add some padding to the bounding box for better visualization
padding <- 100  # Adjust as needed for your visualization
zoom_bbox <- house_1_bbox + c(-padding, -padding, padding, padding)

png("Figure_5.png", width = 6, height = 3, units = "in", res = 300)

# Plot the zoomed-in map
plot(nlcd_data,
     xlim = c(zoom_bbox["xmin"], zoom_bbox["xmax"]), # Set X-axis limits for zoom
     ylim = c(zoom_bbox["ymin"], zoom_bbox["ymax"]) # Set Y-axis limits for zoom
)

# Add the house_1_buffer on top with the cyan fill and blue border
plot(st_geometry(house_1_buffer),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3), # Cyan fill with 30% transparency
     border = "blue", # Blue border
     lwd = 1) # Line width for the border

dev.off()
```

Visually it matches our expectations, but you can also see the detail limitation of the NLCD data we're using. Our recommendation is to use higher boundary distances for broader applications.

Here we add a 300M for single buffer comparison
```{r}
  
# Repeat the prior process.
houses_buffer300m_example <- st_buffer(ames_sf_transformed, dist = 300)

land_cover_raw <- exact_extract(nlcd_data, houses_buffer300m_example)

# Convert the list of extracted values to a data frame
land_cover_df <- land_cover_raw %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq))

# No need to recreate nlcd mapping

land_cover_df <- land_cover_df %>%
  mutate(
    Land_Cover_Type = rename_map[as.character(Var1)]
  )

# Check for any unmapped NLCD codes
unmapped_values <- land_cover_df %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values) > 0) {
  message("Unmapped land cover types detected: ", 
          paste(unique(unmapped_values$Var1), collapse = ", "))
}

# Again optionally, filter out rows with NA land cover types to clean the data
land_cover_df_300m_example <- land_cover_df %>% filter(!is.na(Land_Cover_Type))

# Pivot to wide format and convert Buffer_ID to integer
land_cover_wide <- land_cover_df_300m_example %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(
    names_from = Land_Cover_Type,
    values_from = Proportion,
    values_fill = 0 
  ) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))
```

```{r, fig.width=14, fig.height=8}
# Filter for the buffer and overlay it on the map
house_1_buffer300m <- houses_buffer300m_example %>% filter(X == 1)  # Select buffer for house 1 using X == 1

# Plot the buffer and NLCD data
plot(nlcd_data,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

plot(st_geometry(house_1_buffer300m),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)

# Get the bounding box of the buffer for house 1
house_1_bbox300m <- st_bbox(house_1_buffer300m)

# Add some padding to the bounding box for better visualization
padding <- 300  # Adjust as needed for your visualization
zoom_bbox <- house_1_bbox300m + c(-padding, -padding, padding, padding)


png("Figure_6.png", width = 14, height = 8, units = "in", res = 300)
# Plot the zoomed-in map
plot(nlcd_data,
     xlim = c(zoom_bbox["xmin"], zoom_bbox["xmax"]), # Set X-axis limits for zoom
     ylim = c(zoom_bbox["ymin"], zoom_bbox["ymax"]) # Set Y-axis limits for zoom
)


# Add the house_1_buffer300m on top with the cyan fill and blue border
plot(st_geometry(house_1_buffer300m),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)
dev.off()
```

Let's visualize the buffer once again
```{r}
# Join the wide land cover data to ames_sf_transformed using Buffer_ID
ames_sf_300m_example <- ames_sf_transformed %>%
  left_join(land_cover_wide, by = c("X" = "Buffer_ID"))  # Merge land cover data into spatial dataframe

# Filter for the first house (X == 1)
house_1_stats300m <- land_cover_wide %>%
  filter(Buffer_ID == 1) %>%  # Filter to the first buffer (ID 1)
  pivot_longer(
    cols = -Buffer_ID,  # Select all columns except Buffer_ID for reshaping
    names_to = "Land_Cover_Type",  # Create a column for land cover types
    values_to = "Proportion"  # Create a column for proportions
  )

# View the stats
house_1_stats300m

# Bar plot of proportions for house 1
ggplot(house_1_stats300m, aes(x = Land_Cover_Type, y = Proportion, fill = Land_Cover_Type)) +
  geom_bar(stat = "identity") +  # Create a bar chart where bar heights represent proportions
  scale_fill_viridis_d(name = "Land Cover Type", guide = "none") +  # Discrete color scale
  labs(
    x = "Land Cover Type",  # X-axis label
    y = "Proportion"  # Y-axis label
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```

We now pick up on land cover types that weren't within the 100m buffer




Let's compare the differences in the data
```{r}
# Add buffer labels
house_1_stats <- house_1_stats %>%
  mutate(Buffer = "100m")

house_1_stats300m <- house_1_stats300m %>%
  mutate(Buffer = "300m")

# Combine both datasets
combined_stats <- bind_rows(house_1_stats, house_1_stats300m)

# Combined bar chart
ggplot(combined_stats, aes(x = Land_Cover_Type, y = Proportion, fill = Buffer)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d(name = "Buffer Distance") +
  labs(
    x = "Land Cover Type",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("Figure_7.png", width = 10, height = 6, dpi = 300)
```



From here we can conduct exploratory data analysis on our resulting land cover types from the 100m example; we calculate mean, standard deviation, and other statistics for each land cover type across all buffers.

We can then analyze spatial relationships between houses, neighborhoods, and land cover. Choose to visualize these relationships with high-quality plots and maps. Or integrate multiple data sources to derive meaningful insights about the Ames area.
```{r}
# Calculate summary statistics for the entire dataset
land_cover_summary <- land_cover_df_100m_example %>%
  group_by(Land_Cover_Type) %>%
  summarise(
    Mean_Proportion = mean(Proportion),  # Average proportion per buffer
    SD_Proportion = sd(Proportion),  # Standard deviation of proportion
    Min_Proportion = min(Proportion),  # Minimum proportion
    Max_Proportion = max(Proportion)   # Maximum proportion
  ) %>%
  arrange(desc(Mean_Proportion)) #Arranges the land cover from highest average coverage around each home to the lowest

# Print the summary statistics
print(land_cover_summary)
```

From here we can tell that Low Intensity developed is the highest average proportion whereas barren land is the lowest at just 2.7%


```{r}
# Bar plot of the proportion of each land cover type across all buffers
ggplot(land_cover_df_100m_example, aes(x = Land_Cover_Type, y = Proportion, fill = Land_Cover_Type)) +
  geom_bar(stat = "summary", fun = "mean", show.legend = FALSE) +
  theme_minimal() +
  coord_flip() +  # Flip axis for better readability
  labs(
    x = "Land Cover Type",
    y = "Average Proportion"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
# Plot a boxplot of proportions by land cover type
ggplot(land_cover_df_100m_example, aes(x = Land_Cover_Type, y = Proportion, fill = Land_Cover_Type)) +
  geom_boxplot() +
  labs(
       x = "Land Cover Type",
       y = "Proportion") +
  scale_fill_viridis_d(name = "Land Cover Type") +
  theme_minimal() +
  theme(
    legend.position = "none",  # Remove legend
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  )
```


```{r}
# Frequency plot for a specific land cover type (e.g., "Developed, High Intensity")
ggplot(land_cover_df_100m_example %>% filter(Land_Cover_Type == "Developed, High Intensity"), 
       aes(x = Proportion)) +
  geom_histogram(fill = "steelblue", alpha = 0.6) +
  theme_minimal() +
  labs(
    x = "Proportion",
    y = "Frequency"
  )

```




Section 2. This section builds upon section 1 to create the Dataset with varied buffer lengths and Separate NLCD coverage years. Unfortunately the R package only has access to the legacy version and annual (Released on late October of 2024) is unavailable. The legacy version does have the 2006 2008 and 2011. Where we will apply 2006 -> 2006 & 2007, 2008 -> 2008 & 2009, and 2011 -> to 2010. Barriers are still from 2011 due to the tigris package being updated with the 2010 census, due to the amount of coverage possible barrier mismatch is not an issue.

###############Creating Spatial Dataset #####################

First we split the original dataset into 3 parts based on year
```{r}
ames0607 <- ames %>%
  filter(Year_Sold == 2006 | Year_Sold == 2007)

ames0809 <- ames %>%
  filter(Year_Sold == 2008 | Year_Sold == 2009)

ames10 <- ames %>%
  filter(Year_Sold == 2010)

```

Then we pull the county data for years 2006 2008 and 2011.
```{r}

# Retrieve NLCD (National Land Cover Database) data for Ames, again using the Scaled version due to the largest buffer we'll be using
nlcd_data_0607 <- 
  FedData::get_nlcd(
    template = specific_city_scaled,    
    year = 2006,         
    label = "Ames06"     
  )

nlcd_data_0809 <- 
  FedData::get_nlcd(
    template = specific_city_scaled,    
    year = 2008,         
    label = "Ames08"     
  )

nlcd_data_10 <- 
  FedData::get_nlcd(
    template = specific_city_scaled,    
    year = 2011,         
    label = "Ames10"     
  )


```

Transform and view differences
```{r, fig.width=14, fig.height=8}

# Plot NLCD raster
raster::plot(nlcd_data_0607)


# Plot NLCD raster
raster::plot(nlcd_data_0809)

# Plot NLCD raster
raster::plot(nlcd_data_10)


```
We can see that there is some variation happening throughout the time period



Turn into shape files
```{r}
ames_sf0607 <- st_as_sf(ames0607, coords = c('Longitude', 'Latitude'))
ames_sf0809 <- st_as_sf(ames0809, coords = c('Longitude', 'Latitude'))
ames_sf10 <- st_as_sf(ames10, coords = c('Longitude', 'Latitude'))
```

Declare starting CRS
```{r}
# Assign a CRS (WGS84, EPSG:4326)
st_crs(ames_sf0607) <- 4326
st_crs(ames_sf0809) <- 4326
st_crs(ames_sf10) <- 4326
```

```{r}
# Transform neighborhoods to match the NLCD CRS (ESPG:5070)
ames_sf_transformed0607 <- st_transform(ames_sf0607, st_crs(nlcd_data_0607)) 
ames_sf_transformed0809 <- st_transform(ames_sf0809, st_crs(nlcd_data_0809))
ames_sf_transformed10 <- st_transform(ames_sf10, st_crs(nlcd_data_10)) 
```

create 300m 600m and 1000m buffers and extract. Code is mostly repetition of previous work.

Check work with plots
```{r, fig.width=14, fig.height=8}

# Plot the NLCD raster for each time period but zoomed to original specific_city extent
plot(nlcd_data_0607,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Re add dots
plot(st_geometry(ames_sf_transformed0607),
     add = TRUE,
     pch = 21,                       
     bg = "cyan",                    
     cex = 1)  

# Plot the NLCD raster for each time period but zoomed to original specific_city extent
plot(nlcd_data_0809,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Re add dots
plot(st_geometry(ames_sf_transformed0809),
     add = TRUE,
     pch = 21,                       
     bg = "cyan",                    
     cex = 1)

# Plot the NLCD raster for each time period but zoomed to original specific_city extent
plot(nlcd_data_10,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Re add dots
plot(st_geometry(ames_sf_transformed10),
     add = TRUE,
     pch = 21,                       
     bg = "cyan",                    
     cex = 1)  

```

300m

One time period at a time an then binding them together
```{r, fig.width=14, fig.height=8}

# This creates a buffer zone of 300 meters around each house feature in the spatial dataset.
houses_buffer300 <- st_buffer(ames_sf_transformed0607, dist = 300)

# Extract land cover data values within the new buffer
# exact_extract retrieves raster values that intersect with each buffer zone
land_cover_raw300 <- exact_extract(nlcd_data_0607, houses_buffer300)
names(land_cover_raw300) <- ames_sf0607$X  # Assign original IDs as names

#Save Buffers for later for file output
houses_buffer300_0607 <- houses_buffer300 %>%
  select(X, geometry)

# Plot the NLCD raster
plot(nlcd_data_0607,
     xlim = c(bbox["xmin"], bbox["xmax"]),
     ylim = c(bbox["ymin"], bbox["ymax"]))

# Add semi-transparent red buffers on top
plot(st_geometry(houses_buffer300),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)



```

Transform the output into readable data
```{r}

# Convert the list of extracted values to a data frame
land_cover_df300 <- land_cover_raw300 %>%
  # Apply a function to each element in the list, converting it into a data frame with frequency counts
  lapply(function(x) as.data.frame(table(x$value))) %>%
  # Combine all individual data frames into one, adding an ID column for each buffer
  bind_rows(.id = "Buffer_ID") %>%
  # Group by Buffer_ID for calculations within each buffer
  group_by(Buffer_ID) %>%
  # Calculate the proportion of each land cover type within each buffer
  mutate(Proportion = Freq / sum(Freq))

#Rename map already created, skipping this step

# Add a new column to the data frame for land cover types
land_cover_df300 <- land_cover_df300 %>%
  mutate(
    Land_Cover_Type = rename_map[as.character(Var1)]  # Map codes to land cover names
  )

# Check for any unmapped NLCD codes
unmapped_values300 <- land_cover_df300 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values300) > 0) {
  message("Unmapped land cover types detected: ", 
          paste(unique(unmapped_values300$Var1), collapse = ", "))
}

# Optionally, filter out rows with NA land cover types to clean the data
land_cover_df300 <- land_cover_df300 %>% filter(!is.na(Land_Cover_Type))

# Summarize the cleaned and mapped data
summary(land_cover_df300)

```





```{r}
# Pivot land_cover_df to wide format and convert Buffer_ID to integer
land_cover_wide300 <- land_cover_df300 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%  # Select relevant columns for pivoting
  pivot_wider(
    names_from = Land_Cover_Type,  # Column names will be land cover types
    values_from = Proportion,  # Cell values will be the proportion of each land cover type
    values_fill = 0  # Fill missing land cover types with 0
  ) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))  # Convert Buffer_ID to integer for joining compatibility

# Check the updated wide-format dataframe to confirm the structure
head(land_cover_wide300)

```

```{r}
#Save and repeat
Store300mwide <- land_cover_wide300
```
Create the other 2 time periods and bind them


The following chunk is pure repetition of the prior work
```{r}
# 2008–2009
houses_buffer300 <- st_buffer(ames_sf_transformed0809, dist = 300)

land_cover_raw300 <- exact_extract(nlcd_data_0809, houses_buffer300)
names(land_cover_raw300) <- ames_sf0809$X

#Save Buffers for later for file output
houses_buffer300_0809 <- houses_buffer300 %>%
  select(X, geometry)

# Transform the output into readable data
land_cover_df300 <- land_cover_raw300 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values300 <- land_cover_df300 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values300) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values300$Var1), collapse = ", "))
}
summary(land_cover_df300)

# Convert to wide df format
land_cover_wide300 <- land_cover_df300 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save and repeat
Store300mwide <- rbind(Store300mwide, land_cover_wide300)

# 2010
houses_buffer300 <- st_buffer(ames_sf_transformed10, dist = 300)

land_cover_raw300 <- exact_extract(nlcd_data_10, houses_buffer300)
names(land_cover_raw300) <- ames_sf10$X

#Save Buffers for later for file output
houses_buffer300_10 <- houses_buffer300 %>%
  select(X, geometry)

# Transform the output into readable data
land_cover_df300 <- land_cover_raw300 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values300 <- land_cover_df300 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values300) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values300$Var1), collapse = ", "))
}
summary(land_cover_df300)

# Convert to wide df format
land_cover_wide300 <- land_cover_df300 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save
Store300mwide <- rbind(Store300mwide, land_cover_wide300)

```





```{r}

# Join the wide land cover data using Buffer_ID
ames_spatialized_300m <- ames_sf %>%
  left_join(Store300mwide, by = c("X" = "Buffer_ID"))  # Merge land cover data into spatial dataframe
ames_spatialized_300m_tib <- as_tibble(ames_spatialized_300m)  # Convert the spatial dataframe into a tibble for easier inspection, this is purely for readability and not necessary

# View the updated spatial dataframe with land cover proportions added
head(ames_spatialized_300m_tib)

```
















600m
```{r, fig.width=14, fig.height=8}

# This creates a buffer zone of 600 meters around each house feature in the spatial dataset.
houses_buffer600 <- st_buffer(ames_sf_transformed0607, dist = 600)

# Extract land cover data values within the new buffer
# exact_extract retrieves raster values that intersect with each buffer zone
land_cover_raw600 <- exact_extract(nlcd_data_0607, houses_buffer600)
names(land_cover_raw600) <- ames_sf0607$X  # Assign original IDs as names

#Save Buffers for later for file output
houses_buffer600_0607 <- houses_buffer600 %>%
  select(X, geometry)
# Plot the NLCD raster (Altered to expand view for visual coverage)
plot(nlcd_data_0607,
     xlim = c(bbox["xmin"] - 0.05 * (bbox["xmax"] - bbox["xmin"]),
              bbox["xmax"] + 0.05 * (bbox["xmax"] - bbox["xmin"])),
     ylim = c(bbox["ymin"] - 0.05 * (bbox["ymax"] - bbox["ymin"]),
              bbox["ymax"] + 0.05 * (bbox["ymax"] - bbox["ymin"])))

# Add semi-transparent red buffers on top
plot(st_geometry(houses_buffer600),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)


```

Transform the output into readable data
```{r}

# Convert the list of extracted values to a data frame
land_cover_df600 <- land_cover_raw600 %>%
  # Apply a function to each element in the list, converting it into a data frame with frequency counts
  lapply(function(x) as.data.frame(table(x$value))) %>%
  # Combine all individual data frames into one, adding an ID column for each buffer
  bind_rows(.id = "Buffer_ID") %>%
  # Group by Buffer_ID for calculations within each buffer
  group_by(Buffer_ID) %>%
  # Calculate the proportion of each land cover type within each buffer
  mutate(Proportion = Freq / sum(Freq))

#Rename map already created, skipping this step

# Add a new column to the data frame for land cover types
land_cover_df600 <- land_cover_df600 %>%
  mutate(
    Land_Cover_Type = rename_map[as.character(Var1)]  # Map codes to land cover names
  )

# Check for any unmapped NLCD codes
unmapped_values600 <- land_cover_df600 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values600) > 0) {
  message("Unmapped land cover types detected: ", 
          paste(unique(unmapped_values300$Var1), collapse = ", "))
}

# Optionally, filter out rows with NA land cover types to clean the data
land_cover_df600 <- land_cover_df600 %>% filter(!is.na(Land_Cover_Type))

# Summarize the cleaned and mapped data
summary(land_cover_df600)

```

Transform into wide
```{r}
# Pivot land_cover_df to wide format and convert Buffer_ID to integer
land_cover_wide600 <- land_cover_df600 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%  # Select relevant columns for pivoting
  pivot_wider(
    names_from = Land_Cover_Type,  # Column names will be land cover types
    values_from = Proportion,  # Cell values will be the proportion of each land cover type
    values_fill = 0  # Fill missing land cover types with 0
  ) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))  # Convert Buffer_ID to integer for joining compatibility

# Check the updated wide-format dataframe to confirm the structure
head(land_cover_wide600)

```

```{r}
#Save and repeat
Store600mwide <- land_cover_wide600
```
Create the other 2 time periods and bind them

```{r}
# 2008–2009
houses_buffer600 <- st_buffer(ames_sf_transformed0809, dist = 600)

land_cover_raw600 <- exact_extract(nlcd_data_0809, houses_buffer600)
names(land_cover_raw600) <- ames_sf0809$X

#Save Buffers for later for file output
houses_buffer600_0809 <- houses_buffer600 %>%
  select(X, geometry)


# Transform the output into readable data
land_cover_df600 <- land_cover_raw600 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values600 <- land_cover_df600 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values600) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values600$Var1), collapse = ", "))
}
summary(land_cover_df600)

# Convert to wide df format
land_cover_wide600 <- land_cover_df600 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save and repeat
Store600mwide <- rbind(Store600mwide, land_cover_wide600)

# 2010
houses_buffer600 <- st_buffer(ames_sf_transformed10, dist = 600)

land_cover_raw600 <- exact_extract(nlcd_data_10, houses_buffer600)
names(land_cover_raw600) <- ames_sf10$X

#Save Buffers for later for file output
houses_buffer600_10 <- houses_buffer600 %>%
  select(X, geometry)

# Transform the output into readable data
land_cover_df600 <- land_cover_raw600 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values600 <- land_cover_df600 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values600) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values600$Var1), collapse = ", "))
}
summary(land_cover_df600)

# Convert to wide df format
land_cover_wide600 <- land_cover_df600 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save
Store600mwide <- rbind(Store600mwide, land_cover_wide600)
```


```{r}

# Join the wide land cover data using Buffer_ID
ames_spatialized_600m <- ames_sf %>%
  left_join(Store600mwide, by = c("X" = "Buffer_ID"))  # Merge land cover data into spatial dataframe
ames_spatialized_600m_tib <- as_tibble(ames_spatialized_600m)  # Convert the spatial dataframe into a tibble for easier inspection, this is purely for readability and not necessary

# View the updated spatial dataframe with land cover proportions added
head(ames_spatialized_600m_tib)


```


One last time

1000m
```{r, fig.width=14, fig.height=8}

# This creates a buffer zone of 1000 meters around each house feature in the spatial dataset.
houses_buffer1000 <- st_buffer(ames_sf_transformed0607, dist = 1000)

# Extract land cover data values within the new buffer
# exact_extract retrieves raster values that intersect with each buffer zone
land_cover_raw1000 <- exact_extract(nlcd_data_0607, houses_buffer1000)
names(land_cover_raw1000) <- ames_sf0607$X  # Assign original IDs as names

#Save Buffers for later for file output
houses_buffer1000_0607 <- houses_buffer1000 %>%
  select(X, geometry)

# Plot the NLCD raster (Altered to expand view for visual coverage)
plot(nlcd_data_0607,
     xlim = c(bbox["xmin"] - 0.1 * (bbox["xmax"] - bbox["xmin"]),
              bbox["xmax"] + 0.1 * (bbox["xmax"] - bbox["xmin"])),
     ylim = c(bbox["ymin"] - 0.1 * (bbox["ymax"] - bbox["ymin"]),
              bbox["ymax"] + 0.1 * (bbox["ymax"] - bbox["ymin"])))

# Add semi-transparent red buffers on top
plot(st_geometry(houses_buffer1000),
     add = TRUE,
     col = adjustcolor("cyan", alpha.f = 0.3),
     border = "blue",
     lwd = 1)


```

Transform the output into readable data
```{r}

# Convert the list of extracted values to a data frame
land_cover_df1000 <- land_cover_raw1000 %>%
  # Apply a function to each element in the list, converting it into a data frame with frequency counts
  lapply(function(x) as.data.frame(table(x$value))) %>%
  # Combine all individual data frames into one, adding an ID column for each buffer
  bind_rows(.id = "Buffer_ID") %>%
  # Group by Buffer_ID for calculations within each buffer
  group_by(Buffer_ID) %>%
  # Calculate the proportion of each land cover type within each buffer
  mutate(Proportion = Freq / sum(Freq))

#Rename map already created, skipping this step

# Add a new column to the data frame for land cover types
land_cover_df1000 <- land_cover_df1000 %>%
  mutate(
    Land_Cover_Type = rename_map[as.character(Var1)]  # Map codes to land cover names
  )

# Check for any unmapped NLCD codes
unmapped_values900 <- land_cover_df600 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values900) > 0) {
  message("Unmapped land cover types detected: ", 
          paste(unique(unmapped_values300$Var1), collapse = ", "))
}

# Optionally, filter out rows with NA land cover types to clean the data
land_cover_df1000 <- land_cover_df1000 %>% filter(!is.na(Land_Cover_Type))

# Summarize the cleaned and mapped data
summary(land_cover_df1000)

```


```{r}
# Pivot land_cover_df to wide format and convert Buffer_ID to integer
land_cover_wide1000 <- land_cover_df1000 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%  # Select relevant columns for pivoting
  pivot_wider(
    names_from = Land_Cover_Type,  # Column names will be land cover types
    values_from = Proportion,  # Cell values will be the proportion of each land cover type
    values_fill = 0  # Fill missing land cover types with 0
  ) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))  # Convert Buffer_ID to integer for joining compatibility

# Check the updated wide-format dataframe to confirm the structure
head(land_cover_wide1000)

```



```{r}
#Save and repeat
Store1000mwide <- land_cover_wide1000
```
Create the other 2 time periods and bind them

```{r}
# 2008–2009
houses_buffer1000 <- st_buffer(ames_sf_transformed0809, dist = 1000)

land_cover_raw1000 <- exact_extract(nlcd_data_0809, houses_buffer1000)
names(land_cover_raw1000) <- ames_sf0809$X


#Save Buffers for later for file output
houses_buffer1000_0809 <- houses_buffer1000 %>%
  select(X, geometry)

# Transform the output into readable data
land_cover_df1000 <- land_cover_raw1000 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values1000 <- land_cover_df1000 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values1000) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values1000$Var1), collapse = ", "))
}
summary(land_cover_df1000)

# Convert to wide df format
land_cover_wide1000 <- land_cover_df1000 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save and repeat
Store1000mwide <- rbind(Store1000mwide, land_cover_wide1000)

# 2010
houses_buffer1000 <- st_buffer(ames_sf_transformed10, dist = 1000)

land_cover_raw1000 <- exact_extract(nlcd_data_10, houses_buffer1000)
names(land_cover_raw1000) <- ames_sf10$X

#Save Buffers for later for file output
houses_buffer1000_10 <- houses_buffer1000 %>%
  select(X, geometry)

# Transform the output into readable data
land_cover_df1000 <- land_cover_raw1000 %>%
  lapply(function(x) as.data.frame(table(x$value))) %>%
  bind_rows(.id = "Buffer_ID") %>%
  group_by(Buffer_ID) %>%
  mutate(Proportion = Freq / sum(Freq)) %>%
  mutate(Land_Cover_Type = rename_map[as.character(Var1)]) %>%
  filter(!is.na(Land_Cover_Type))

unmapped_values1000 <- land_cover_df1000 %>% filter(is.na(Land_Cover_Type))
if (nrow(unmapped_values1000) > 0) {
  message("Unmapped land cover types detected: ", paste(unique(unmapped_values1000$Var1), collapse = ", "))
}
summary(land_cover_df1000)

# Convert to wide df format
land_cover_wide1000 <- land_cover_df1000 %>%
  select(Buffer_ID, Land_Cover_Type, Proportion) %>%
  pivot_wider(names_from = Land_Cover_Type, values_from = Proportion, values_fill = 0) %>%
  mutate(Buffer_ID = as.integer(Buffer_ID))

# Save
Store1000mwide <- rbind(Store1000mwide, land_cover_wide1000)
```



```{r}

# Join the wide land cover data using Buffer_ID
ames_spatialized_1000m <- ames_sf %>%
  left_join(Store1000mwide, by = c("X" = "Buffer_ID"))  # Merge land cover data into spatial dataframe
ames_spatialized_1000m_tib <- as_tibble(ames_spatialized_1000m)  # Convert the spatial dataframe into a tibble for easier inspection, this is purely for readability and not necessary

# View the updated spatial dataframe with land cover proportions added
head(ames_spatialized_1000m)

```

Combine the new data into one
```{r}
# Function to rename land cover columns
rename_land_cover <- function(data, distance) {
  data %>%
    rename_with(~ paste0(., " ", distance), 
                matches("^(Open Water|Perennial Ice/Snow|Developed, Open Space|Developed, Low Intensity|Developed, Medium Intensity|Developed, High Intensity|Barren Land|Deciduous Forest|Evergreen Forest|Mixed Forest|Dwarf Scrub|Shrub/Scrub|Grassland/Herbaceous|Sedge/Herbaceous|Lichens|Moss|Pasture/Hay|Cultivated Crops|Woody Wetlands|Emergent Herbaceous Wetlands)$"))
}

# Rename land cover columns
houses_300m <- rename_land_cover(ames_spatialized_300m, "300m")
houses_600m <- rename_land_cover(ames_spatialized_600m, "600m")
houses_1000m <- rename_land_cover(ames_spatialized_1000m, "1000m")

# Combine datasets (assuming they have the same number of rows and identical other columns)
combined_houses <- bind_cols(
  houses_300m,
  select(houses_600m, matches("^(Open Water|Perennial Ice/Snow|Developed, Open Space|Developed, Low Intensity|Developed, Medium Intensity|Developed, High Intensity|Barren Land|Deciduous Forest|Evergreen Forest|Mixed Forest|Dwarf Scrub|Shrub/Scrub|Grassland/Herbaceous|Sedge/Herbaceous|Lichens|Moss|Pasture/Hay|Cultivated Crops|Woody Wetlands|Emergent Herbaceous Wetlands) 600m$")),
  select(houses_1000m, matches("^(Open Water|Perennial Ice/Snow|Developed, Open Space|Developed, Low Intensity|Developed, Medium Intensity|Developed, High Intensity|Barren Land|Deciduous Forest|Evergreen Forest|Mixed Forest|Dwarf Scrub|Shrub/Scrub|Grassland/Herbaceous|Sedge/Herbaceous|Lichens|Moss|Pasture/Hay|Cultivated Crops|Woody Wetlands|Emergent Herbaceous Wetlands) 1000m$"))
)

# View combined dataset
head(combined_houses)

```







Output Spacialized Ames File
```{r}
# Save table as a CSV file
combined_houses %>%
  write_csv("ames_spatialized.csv")

```

Save and Output Buffers
```{r}
#First Bind matching buffers
# Combine all 300m buffers with period (This is for the landscape metrics)
buffer_300m <- bind_rows(
  mutate(houses_buffer300_0607, period = "0607"),
  mutate(houses_buffer300_0809, period = "0809"),
  mutate(houses_buffer300_10,   period = "10")
)

# Combine all 600m buffers with period
buffer_600m <- bind_rows(
  mutate(houses_buffer600_0607, period = "0607"),
  mutate(houses_buffer600_0809, period = "0809"),
  mutate(houses_buffer600_10,   period = "10")
)

# Combine all 1000m buffers with period
buffer_1000m <- bind_rows(
  mutate(houses_buffer1000_0607, period = "0607"),
  mutate(houses_buffer1000_0809, period = "0809"),
  mutate(houses_buffer1000_10,   period = "10")
)
```


```{r}

#Then output files
st_write(buffer_300m,  "ames_buffer_300m.shp",  delete_dsn = TRUE)
st_write(buffer_600m,  "ames_buffer_600m.shp",  delete_dsn = TRUE)
st_write(buffer_1000m, "ames_buffer_1000m.shp", delete_dsn = TRUE)
```

Adding Landscape metrics
```{r}
library("landscapemetrics")
```



Aggregate index
```{r}
#prepare buffers for metrics, cleaner names are used here
buffers_0607 <- buffer_1000m %>% filter(period == "0607")
buffers_0809 <- buffer_1000m %>% filter(period == "0809")
buffers_10   <- buffer_1000m %>% filter(period == "10")
```


```{r}
#Changing from a spatraster to a regular raster for compatability
nlcd_0607_rast <- raster(nlcd_data_0607)
nlcd_0809_rast <- raster(nlcd_data_0809)
nlcd_10_rast   <- raster(nlcd_data_10)
```

This first pass will be very verbose but it applies conceptually to all 3 landscape metric functions we'll do
```{r}
# Function to calculate Aggregation metrics (AI) for selected land cover classes
calculate_aggregation_index <- function(raster_input, buffer_sf) {
  # Preserve the original unique IDs from buffer_sf by copying column 'X' into 'orig_X'
  # This is necessary because sample_lsm will reset buffer IDs internally
  buffer_sf <- buffer_sf %>%
    mutate(orig_X = X)
  
  # Calculate the Aggregation Index (AI) metric for each land cover class
  # within each buffer polygon using sample_lsm
  result <- sample_lsm(
    landscape = raster_input,   # The NLCD raster data
    y = buffer_sf,              # The spatial polygons (buffers) to analyze
    level = "class",            # Calculate metrics at the land cover class level
    metric = "ai",              # The specific metric: Aggregation Index
    directions = 8             # Consider 8 neighboring cells for adjacency (queen’s case)
  )
  
  # sample_lsm returns results with plot_id assigned as 1 to number of buffers processed.
  # To map results back to original buffer IDs:
  result <- result %>%
    mutate(plot_id = as.integer(plot_id)) %>%  # Ensure plot_id is integer for join
    # Join with original buffer_sf using plot_id as row number to preserve IDs
    left_join(buffer_sf %>% mutate(plot_id = row_number()), by = "plot_id") %>%
    # Create a descriptive metric label by combining 'ai_' with land cover class code
    mutate(metric_label = paste0("ai_", class)) %>%
    # Select only the original ID, the metric label, and the metric value columns
    select(orig_X, metric_label, value) %>%
    # Rename 'orig_X' back to 'X' for consistency with original data
    rename(X = orig_X)
  
  # Return the final cleaned data frame
  return(result)
}
```



```{r}
# Calculate AI metrics for each time period
ai_0607 <- calculate_aggregation_index(nlcd_0607_rast, buffers_0607)
ai_0809 <- calculate_aggregation_index(nlcd_0809_rast, buffers_0809)
ai_10   <- calculate_aggregation_index(nlcd_10_rast,   buffers_10)

```

```{r}
#Bind the parts together
ai_all <- bind_rows(ai_0607, ai_0809, ai_10)
```


```{r}
ai_wide <- ai_all %>%
  pivot_wider(names_from = metric_label, values_from = value) %>%
  arrange(X)
```

```{r}
#Clean up the names a bit
colnames(ai_wide) <- sub("^ai_", "", colnames(ai_wide))
```

```{r}
#rename using prior nlcd mapping
colnames(ai_wide) <- sapply(colnames(ai_wide), function(col) {
  match_row <- nlcd_mapping$Land_Cover_Type[nlcd_mapping$Column == col]
  if (length(match_row) == 1) match_row else col
})
```

```{r}
#Re add aggregate index indicator for clarity
colnames(ai_wide)[colnames(ai_wide) != "X"] <- paste0("ai ", colnames(ai_wide)[colnames(ai_wide) != "X"])
```

Attach
```{r}
# Attach to the housing dataset
combined_houses <- combined_houses %>%
  left_join(ai_wide, by = "X")
```

Add Shannon and Simpson diversity metrics


```{r}
# Function to calculate landscape-level diversity metrics + patch counts (total) in each buffer
calculate_diversity_metrics <- function(raster_input, buffer_sf) {
  buffer_sf <- buffer_sf %>% mutate(orig_X = X)
  
  result <- sample_lsm(
    landscape = raster_input,
    y = buffer_sf,
    level = "landscape",
    metric = c("shdi", "shei", "sidi", "siei", "np"),
    directions = 8
  )
  
  result <- result %>%
    mutate(plot_id = as.integer(plot_id)) %>%
    left_join(buffer_sf %>% mutate(plot_id = row_number()), by = "plot_id") %>%
    select(orig_X, metric, value) %>%
    rename(X = orig_X, metric_label = metric)
  
  return(result)
}

```

```{r}
div_0607 <- calculate_diversity_metrics(nlcd_0607_rast, buffers_0607)
div_0809 <- calculate_diversity_metrics(nlcd_0809_rast, buffers_0809)
div_10   <- calculate_diversity_metrics(nlcd_10_rast,   buffers_10)

```


```{r}
div_all <- bind_rows(div_0607, div_0809, div_10)

div_wide <- div_all %>%
  pivot_wider(names_from = metric_label, values_from = value) %>%
  arrange(X)

```

```{r}
combined_houses <- combined_houses %>%
  left_join(div_wide, by = "X")

```

patch count at the class level (Instead of total patch count it's broken by number of each patch type)
```{r}
# Function to calculate class-level patch count (NP)
calculate_class_patchcount <- function(raster_input, buffer_sf) {
  buffer_sf <- buffer_sf %>%
    mutate(orig_X = X)
  
  result <- sample_lsm(
    landscape = raster_input,
    y = buffer_sf,
    level = "class",
    metric = "np",  # class-level patch count
    directions = 8
  )
  
  result <- result %>%
    mutate(plot_id = as.integer(plot_id)) %>%
    left_join(buffer_sf %>% mutate(plot_id = row_number()), by = "plot_id") %>%
    mutate(metric_label = paste0("np_", class)) %>%
    select(orig_X, metric_label, value) %>%
    rename(X = orig_X)
  
  return(result)
}
```

```{r}
np_class_0607 <- calculate_class_patchcount(nlcd_0607_rast, buffers_0607)
np_class_0809 <- calculate_class_patchcount(nlcd_0809_rast, buffers_0809)
np_class_10   <- calculate_class_patchcount(nlcd_10_rast,   buffers_10)

```

```{r}
np_class_all <- bind_rows(np_class_0607, np_class_0809, np_class_10)

np_class_wide <- np_class_all %>%
  pivot_wider(names_from = metric_label, values_from = value) %>%
  arrange(X)

```

```{r}
# Remove "np_" prefix
colnames(np_class_wide) <- sub("^np_", "", colnames(np_class_wide))

# Apply land cover type labels using nlcd_mapping
colnames(np_class_wide) <- sapply(colnames(np_class_wide), function(col) {
  match_row <- nlcd_mapping$Land_Cover_Type[nlcd_mapping$Column == col]
  if (length(match_row) == 1) match_row else col
})

# Add "patch count" prefix (with space)
colnames(np_class_wide)[colnames(np_class_wide) != "X"] <- paste0("np ", colnames(np_class_wide)[colnames(np_class_wide) != "X"])

```

```{r}
combined_houses <- combined_houses %>%
  left_join(np_class_wide, by = "X")
```


Write second csv
```{r}
# Save table as a CSV file
combined_houses %>%
  write_csv("ames_spatialized_indexed.csv")

```

Make summary statistics
```{r}
#install.packages("psych")
library(psych)

# Select variables of interest (AKA all new variables)
all_cols <- colnames(combined_houses)

# Land cover buffers ending with 300m, 600m, or 1000m
landcover_buffer_vars <- grep(" (300m|600m|1000m)$", all_cols, value = TRUE)

# Variables starting with "ai "
ai_vars <- grep("^ai ", all_cols, value = TRUE)

# Variables starting with "np"
np_vars <- grep("^np", all_cols, value = TRUE)

# Exact diversity index names
diversity_indices <- c("shdi", "shei", "sidi", "siei")
diversity_vars <- intersect(diversity_indices, all_cols)

# Combine all
vars_of_interest <- c(landcover_buffer_vars, ai_vars, np_vars, diversity_vars)

vars_of_interest

# Generate summary stats
summary_stats <- describe(combined_houses[vars_of_interest])[ , c("mean", "min", "max", "n")]


# Save to CSV
write.csv(summary_stats, "summary_stats.csv", row.names = TRUE)
```
